---
title: "DATA 607 Assignment Three 9-12-21"
author: "Joseph Foy"
date: "9/12/2021"
output:
  html_document: default
  number_sections: true
  pdf_document: default
  df_print: paged
  word_document: default
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Question One

### Overview

#### At FiveThirtyEight, the assigned article is titled, "The Economic Guide To Picking A College Major" at https://fivethirtyeight.com/features/the-economic-guide-to-picking-a-college-major/.  This focus of this article to increase the chances of economic success when selecting a major.  This dataset contains a majors-list file originating from the Bureau of Labor Statistics 2010-2012 Public Use Microdata Series.  The variables include field of degree (FOD1P), A Major, and Major Category.  The file is sorted by Major Category.  Major categories are from Carnevale et al, "What's It Worth?: The Economic Value of College Majors." Georgetown University Center on Education and the Workforce, 2011. http://cew.georgetown.edu/whatsitworth.

#### First, import the dataset. 

```{r load libraries and data}
library(tidyverse)
library(dplyr)
library(ggplot2)
College_Majors  = read_csv("https://raw.githubusercontent.com/fivethirtyeight/data/master/college-majors/majors-list.csv",show_col_types = FALSE)
```

#### View the variable names with the "str" command.

```{r View Variables}
str(College_Majors)
```

#### View the dataframe.

```{r}
head(College_Majors)
```

#### Subset the majors with DATA in the name and its location in the dataframe.
```{r}
str_subset(College_Majors$Major,'DATA')
grep('DATA', College_Majors$Major)
```

#### Subset the majors with STATISTICS in the name and its location in the dataframe.
```{r}
str_subset(College_Majors$Major,'STATISTICS')
grep('STATISTICS', College_Majors$Major)
```

## Question Two

#### Write code that transforms the data below:

[1] "bell pepper"  "bilberry"     "blackberry"   "blood orange"

[5] "blueberry"    "cantaloupe"   "chili pepper" "cloudberry"  

[9] "elderberry"   "lime"         "lychee"       "mulberry"    

[13] "olive"        "salal berry"

Into a format like this:

c("bell pepper", "bilberry", "blackberry", "blood orange", "blueberry", "cantaloupe", "chili pepper", "cloudberry", "elderberry", "lime", "lychee", "mulberry", "olive", "salal berry")

#### Create a list of the character data.

```{r}
fruit <- list("bell pepper", "bilberry", "blackberry", "blood orange", "blueberry", "cantaloupe", "chili pepper", "cloudberry", "elderberry", "lime", "lychee", "mulberry", "olive", "salal berry")
```

#### Transform to a Vector.

```{r}
unlist(fruit)
```

## Question Three

#### Describe, in words, what these expressions will match:

A. (.)\1\1

This example does not have two backslashes.  If it did, this matches any character, then repeat the first character and repeat the first character again. Example: ccc.

B. "(.)(.)\\2\\1"

This matches any character, any character, then repeat the second character, and then repeat the first character. Example: caac.

C. (..)\1

This example does not have two backslashes.  If it did, this matches any two characters and then repeat them. Example: caca.

D. "(.).\\1.\\1"

This matches any character, any character, then repeat the first character, any character, and then repeat the first character. Example: cacbc.

E. "(.)(.)(.).*\\3\\2\\1"

This matches any character at least four times, then repeat the third, second and first characters in that order. Example: abcdddcba.

## Question Four

#### Construct regular expressions to match words that:

A. Start and end with the same character.

"^(.).*\\1$"

The ^ starts the word, the (.) is first character, the . is another character, the * allows for more characters and the \\1& repeats the first character.

B. Contain a repeated pair of letters (e.g. "church" contains "ch" repeated twice.)

"^(..).*\\1$"

The ^ starts the word, the (..) is first two characters, the . is another character, the * allows for more characters and the \\1 repeats the two characters.

C. Contain one letter repeated in at least three places (e.g. "eleven" contains three "e"s.)

"(.).\\1.\\1"

The (.) is first character, the . is another character, the \\1 repeats the first character, the . is another character, and the \\1 repeats the first character again.