---
title: "Examples of tidyverse"
author: "Tom Buonora and ..."
---
title: "Examples of tidyverse"
author: "Tom Buonora and ..."
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tidyverse Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
   warning = FALSE, 
  message = FALSE,
  fig.width = 8,
  fig.height = 8
)
```

```{r}

library(tidyverse)         # ggplot2, dplyr, tidyr, readr, tibble, sringr and more

```

# Tidyverse Vignette

##   readr : read_csv()

<br>
read_csv is part of readr whereas read.csv is base R. Im not sure that read_csv is tidier than read.csv
<br>

```{r readr : read_csv}
CURR_PATH<-str_trim(getwd())

# to do : use the kaggle api
# https://www.kaggle.com/mysarahmadbhat/nyc-traffic-accidents

# nyc_traffic_accidents<-read.csv("NYC Accidents 2020.csv")

nyc_traffic_accidents<-read_csv("https://raw.githubusercontent.com/TheReallyBigApple/CunyAssignments/main/DATA607/NYC%20Accidents%202020.csv",show_col_types = FALSE)
```

##   readr : spec()

the spec function displays the schema for the data frame or tibble

```{r readr : spec}

# the readr spec method shows you the schema
spec(nyc_traffic_accidents)

# this is base R, it removes the spaces from the column names
colnames(nyc_traffic_accidents) <- make.names(colnames(nyc_traffic_accidents))
```

##   dplyr : group_by()    dplyr_summarise()     magrittr : %>%

magrittr provides several operators including the commonly used %>% which pipes results from one function to the next

```{r several}
nyc_traffic_accidents %>%
  group_by(BOROUGH) %>%
  summarise(observations = n(), killed = sum(NUMBER.OF.PERSONS.KILLED, na.rm = TRUE))
```
```{r}
library(leaflet)
```

##   dplyr : select()

In this area, I extended this file by filtering out and mapping people killed in traffic accidents in NYC. In the original file, an "nyc_traffic_accidents" object was created. In the original coding, the authors filtered out the latitude and longitude columns.  However, to map the data, the goecoding data is necessary.  To preserve the original coding to show what the original authors created, I had to create a new object called "map_accidents" and position it before the authors filtered out the latitude and longitude columns in the Rmarkdown file.  

With the dpylr tool, I selected the columns Borough, Latitude, Longitude and Number of Persons Killed. I removed all the NAs and filtered the Number of Persons Killed with a value of greater than zero.  

```{r dplyr select}
map_accidents<-nyc_traffic_accidents %>% 
    select(BOROUGH, LATITUDE, LONGITUDE, NUMBER.OF.PERSONS.KILLED) %>%
    na.omit(nyc_traffic_accidents) %>%
    filter(NUMBER.OF.PERSONS.KILLED > 0) 
map_accidents
```

## Using Leaflet to Map the Data.

In this area, I created a leaflet map.  I set the view in Leaflet so it centers on New York City when the map opens.  I added blue circles to show the location where people were killed. 

```{r} 
# Create Leaflet map centered on Brooklyn.
  leaflet(map_accidents) %>% 
  addProviderTiles("CartoDB") %>%
  setView(-73.95, 40.725, zoom = 10) %>%
  addCircles(lat = ~ LATITUDE, lng = ~ LONGITUDE)
```



##   dplyr : select()

select can select certain columns
select can also omit certain columns. 
Below I omit a sequence of columns starting with LATTITUDE and ending with LOCATION

```{r dplyr select}
nyc_traffic_accidents<-nyc_traffic_accidents %>% 
    select(!(LATITUDE:LOCATION))
```

##   dplyr : muatate()   replace_na()

muate() can create or modify a column
replace_na() can replace all instances of NA with a specified value

```{r replace_na}

nyc_traffic_accidents<-nyc_traffic_accidents %>% dplyr::mutate(BOROUGH = replace_na(BOROUGH, "NYC"))

```

##   ggplot :  geom_histogram() et all.

while the other tidyverse packages infer data parsing, ggplot exists to *display* data
having said that, its robust functionality includes data shaping. The histogram is essentially a group_by()
as is scale_x_dates

```{r ggplot histogram}


ggplot(data = nyc_traffic_accidents, aes(x = CRASH.DATE)) +
geom_histogram() +
theme(axis.text.x=element_text(angle=60, hjust=1)) +
ggtitle("Accidents Per Day") +
scale_x_date(date_breaks = "weeks", date_labels = "%Y-%m-%d") +
xlab("Date") 
```




```

##   dplyr : muatate()   replace_na()

muate() can create or modify a column
replace_na() can replace all instances of NA with a specified value

```{r replace_na}

nyc_traffic_accidents<-nyc_traffic_accidents %>% dplyr::mutate(BOROUGH = replace_na(BOROUGH, "NYC"))

```


##   ggplot :  geom_histogram() et all.

while the other tidyverse packages infer data parsing, ggplot exists to *display* data
having said that, its robust functionality includes data shaping. The histogram is essentially a group_by()
as is scale_x_dates

```{r ggplot histogram}


ggplot(data = nyc_traffic_accidents, aes(x = CRASH.DATE)) +
geom_histogram() +
theme(axis.text.x=element_text(angle=60, hjust=1)) +
ggtitle("Accidents Per Day") +
scale_x_date(date_breaks = "weeks", date_labels = "%Y-%m-%d") +
xlab("Date") 
```


