---
title: "TidyVerse CREATE Assigment 10-24-21"
author: "Joseph Foy"
date: "10/24/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Requirements for TidyVerse CREATE Assignment:

### In this assignment, you’ll practice collaborating around a code project with GitHub.  You could consider our collective work as building out a book of examples on how to use TidyVerse functions.

### GitHub repository:  https://github.com/acatlin/FALL2021TIDYVERSE

### FiveThirtyEight.com datasets.

### Kaggle datasets. 

### Your task here is to Create an Example.  Using one or more TidyVerse packages, and any dataset from fivethirtyeight.com or Kaggle, create a programming sample “vignette” that demonstrates how to use one or more of the capabilities of the selected TidyVerse package with your selected dataset. (25 points)

### Later, you'll be asked to extend an existing vignette.  Using one of your classmate’s examples (as created above), you'll then extend his or her example with additional annotated code. (15 points)

### You should clone the provided repository.  Once you have code to submit, you should make a pull request on the shared repository.  You should also update the README.md file with your example.

### After you’ve created your vignette, please submit your GitHub handle name in the submission link provided below. This will let your instructor know that your work is ready to be peer-graded.

### You should complete your submission on the schedule stated in the course syllabus.


### In this markdown file, the dataset ABNYC2019 located at https://www.kaggle.com/dgomonov/new-york-city-airbnb-open-data is explored.  It is from New York City Airbnb Open Data.  This dataset contains summary information and metrics for listings in New York City. It contains 48,896 rows of data across 16 columns.  This markdown takes the dataset and uses TidyVerse to isolate just Air BnB locations in Brooklyn by Longitude and Latitude and then utilizes the package Leaflet to map all Air BnB locations in Brooklyn. 

#### Load .csv file downloaded and stored in GitHub.

```{r}
csv_url <- "https://raw.githubusercontent.com/professorfoy/DATA-607/main/AB_NYC_2019.csv"
```

#### Load Libraries TidyVerse and Leaflet.  A new object labeled ABnB was created for the dataframe.  The TidyVerse package Readr is used to read the .cvs data via the read.csv command.

```{r}
# load libraries and import data
library(tidyverse)
library(leaflet)
ABnB <- read.csv(csv_url, header = FALSE)
```

#### View the variable names with the "str" command.  Confirm that all 48,896 rows of data was imported. All are character variables.  The latitude and longitude variables will eventually have to be converted to numeric variables to plot locations.  

```{r} 
# View ABnB Variables
str(ABnB)
```
#### For purposes of this markdown file, the dataset contains columns of data that are of no interest.  Only the locations by Borough, latitude and longitude are of interest.  A new object labeled ABnB_Locations was created for the dataframe. The select command from the TidyVerse package Dplyr is used to is applied to isolate these columns.  In the dataset, the first row contains the column names.  The slice command the slice commandremove the first column. 

```{r}
# Isolate locations, latitude and longitude.
ABnB_Locations <- select(ABnB, V5, V7, V8) %>% 
  slice(-c(1))
ABnB_Locations
```
#### The headers for the three columns of interest need to be renamed for a better description.  In NYC, the county names are more colloquially known as Boroughs, and therefore column V5 was renamed as Boroughs.  V7 and V8 were renamed Latitude and Longitude, respectively.  A new object labeled ABnB_Brooklyn was created for the dataframe.  Once renamed, the Borough of Brooklyn was isolated by using the filter command.  The rename and filter commands from the TidyVerse package Dplyr is used to rename colums and isolate Brooklyn within the Borough column.

```{r}
# Filter out locations in Brooklyn.
ABnB_Brooklyn <- rename(ABnB_Locations, Borough= 'V5', Latitude = 'V7', Longitude = 'V8') %>%
  filter(Borough=='Brooklyn')
ABnB_Brooklyn
```
#### To be able to plot locations on a map, the latitude and longitude variables must be converted to numeric variables.  The base R command as.numeric is applied to the latitude and longitude variables and the str command is used to verify they have been converted to numeric values (quotation marks no longer appear around values).

```{r}
ABnB_Brooklyn$Latitude <- as.numeric(ABnB_Brooklyn$Latitude)
ABnB_Brooklyn$Longitude <- as.numeric(ABnB_Brooklyn$Longitude)
str(ABnB_Brooklyn)
```

#### Using Leaflet, a new object labeled ABnB_Map was created for the dataframe. To use Leaflet, the first step is to create an centered area to plot the latitude and longitude coordinates.  The addTiles command creates the Leaflet map and the setView command using latitude and longitude coordinates for the center of Brooklyn (found by trial and error) is used as the center of the map.   The addCircles command was used to plot blue circles at each location using the latitude and longitude data from the ABnB_Brooklyn dataframe.

```{r}
# Create Leaflet map centered on Brooklyn.
ABnB_Map <- ABnB_Brooklyn %>% leaflet() %>%
  addTiles() %>%
  setView(-73.95, 40.635, zoom = 11) %>%
  addCircles(lat = ~ Latitude, lng = ~ Longitude)
ABnB_Map
```

#### As one can visualize, Air BnB locations in Brooklyn are heavily centralized in north to northwest Brooklyn and less sparse in south to southeast Brooklyn.  